
---
title: "STAT340 HW1: Random Variables and R Review"
author: "Keith Levin and Bi Cheng Wu"
date: "September 2021"
output: html_document
---

## Problem 1: Random Variables

### 1) Use R to generate a single Binomial, Geometric, and Poisson random variable.

 You can choose the parameter of the distribution to be anything you like.  

Here are the function names: `rbinom, rgeom, rpois`.
Remember that the "r" stands for "Random".

### 1a) Read the help files for these functions. What are the 8 headings in each of the help files? 

Reading the help files is hard at first, but it is an essential skill.  Have a look at the help files for these functions.  What are the 8 headings?
```{r}
?rbinom
?rgeom
?rpois
```

### 1b) Generate a single copy of each random variable. Once again, you may set parameters however you like.

```{r}

# generate a single Binomial random variable:
rbinom(n = 1, size = 10, prob = .3)

# generate a single Geometric random variable:

# generate a single Poisson random variable:
```


### 1c) Generate ten copies of each random variable.

```{r}
# generate 10 Binomial random variables:

# generate 10 Geometric random variables:
rgeom(n=10, p=0.25)

# generate 10 Poisson random variables:

```


### 2) Making histograms

One way to think about random variables is through their "distribution".
To see what the distributions of different random variables look like, generate 10,000 copies of the random variables above and put them in a histogram:

```{r}
# Exponential:
hist(rexp(n = 10000, lambda=5))
# or 
library(magrittr)
rexp(10000, lambda=1) %>% hist

# Binomial:


# Normal:


```

Repeat the last step, but change the value of the parameter that you put in.  

```{r}

# Exponential:
hist(rexp(n = 10000,size = 1,prob = .3))
# or 
library(magrittr)
rexp(10000,1,.3) %>% hist

# Binomial: (change p, not the size parameter)


# Binomial: (change the size parameter, not p)


# Normal: (change the mean, not the standard deviation)


# Normal: (change the standard deviation, not the mean)


```


### 3) Changing parameters, changing distributions

What happens to the histogram when the parameter gets bigger?  

Answer in words:

1) Exponential: the histogram "shrinks toward 0" as lambda increases.

2) Binomial: ...

3) Normal: ...

### 4) Defining some simple functions in R.

This is not a programming course, but it's important that you become familiar with some basic programming concepts this semester.
Toward that end, let's close by getting some practice with function definition in R.

```{r}
# 4a) Define a function double_x that takes a single argument x and returns twice that argument.
# You may assume that x is numeric, so that 2*x makes sense.
double_x <- function( x ){
  return( 2*x );
}

# 4b) Define a function triple_x that takes a single argument x and returns three times that argument.
# You may assume that x is numeric, so that 3*x makes sense.
triple_x <- function( x ) {
  # TODO: CODE GOES HERE!
}

# 4c) Define a function my_multiply that takes two arguments x and y and returns x*y.
# You may assume that x and y are both numeric so that x*y makes sense.

#TODO: define a function my_multiply( x,y).



# 4d) Define a function my_multiply2 that takes two arguments x and y and returns x*y.
# This function should have a default behavior where x and y both default to 1.
# That is, my_multiply2(x,y) should return x*y, my_multiply2(x) should return x, and my_multiply2() should return 1.

# TODO: define a function my_multiply2(x,y) w

```
